---
title: "Prunus pest life stages report"
author: "Dalon White"
date: "2025-06-18"
output: html_document

params:
  commodity: ['Prunus salicina','Prunus domestica']
  begin_date: 1901-01-01
  end_date: 
---

load packages

```{r}
#install.packages('pacman')
pacman::p_load(odbc, tidyverse, DBI, DT, lubridate)
```

put parameters into global env

```{r get params}
commodity_vec <- c("prunus salicina", "prunus domestica", "prunus sp.")
begin_date <- as.Date('1901-01-01')
end_date <- Sys.Date()
```

No clue why the params in the yaml header isn't being picked up right now, it must just be a temporary glitch:

```{r get params}
commodity_vec <- params$commodity
begin_date <- as.Date(params$begin_date)
end_date <- if(!is.na(params$end_date)){as.Date(params$end_date)} else {
  Sys.Date()
}
```

connect to server

```{r connect to db}
# Connect to ARM DM 2 database connection: #####
db_conn <- dbConnect(odbc::odbc(),
                           .connection_string =
                             "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;        
                        trusted_connection=yes")
```

# ARM data

## The easy way

Use the curated mvw_Diagnostic_Results_Current table

```{r get pest records}

mvw_diagnostic_results <- tbl(
  db_conn,
  sql("select * FROM [APHIS_Imports].[dbo].[mvw_Diagnostic_Results]")
)

df_diagnostic_results <- mvw_diagnostic_results %>% 
  filter(
    INSPECTION_DATE >= begin_date,
    INSPECTION_DATE <= end_date,
    DETERMINATION_TYPE == 'Final ID',
    !is.na(PEST_TAXONOMIC_NAME)
  )  |>   collect() |>
  filter(
    grepl(paste(commodity_vec, collapse = "|"), tolower(COMMODITY_DISPLAY_NAME))
    ) |>
    filter(
      grepl(paste(commodity_vec, collapse = "|"), tolower(COMMODITY_DISPLAY_NAME))
           )

# setwd(here::here('output'))
# write.csv(df_diagnostic_results,
#           paste(
#             'Pest interception records in mvw_Diagnostic_Results on',
#             paste(commodity_vec,collapse = 'OR'), 
#             'between', 
#             begin_date, 
#             end_date, 
#             ".csv",
#           sep = " ")
# )
```

### Also requested adding plant parts

```{r attach plant parts data}
get.commodity = function(connection,
                         IDs
) {
  tbl(connection, sql( "SELECT
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_COMMODITY]")) |>
    # dplyr::select(
    #   ID,
    #   INSPECTION_ID,
    #   PLANT_PARTS
    # ) |> 
    collect() |> 
    rename('COMMODITY_ID' = 'ID') |> 
    filter(COMMODITY_ID %in% IDs)
}

plant_part_records <- get.commodity(db_conn, IDs = df_diagnostic_results |> pull(ID) )

commod_records <- df_diagnostic_results |> 
  left_join(plant_part_records, by = c('ID' = 'COMMODITY_ID'))

setwd(here::here('output'))
write.csv(commod_records,
          paste(
            'Pest interception records in mvw_Diagnostic_Results on',
            paste(commodity_vec,collapse = 'OR'), 
            'between', 
            begin_date, 
            end_date, 
            "in ARM",
            ".csv",
          sep = " ")
)
```

# AQAS data

```{r get records from AQAS}
aqas_f309_intercept <- tbl(
  db_conn,
  sql("select * FROM [PPQ_AQI_AQAS_DW].[DW_AQAS].[PESTID_F309_INTERCEPT]")
)

aqas_f309_genus_records <- aqas_f309_intercept  |>  
  filter(
    CREATE_DT >= begin_date,
    CREATE_DT <= end_date,
    #filter by prunus now just so it's a smaller table (slow internet)
    INSP_GENUS %in% "Prunus"
  )  |>   collect() 

#match by commodity_vec
aqas_f309_species_records <- aqas_f309_genus_records |>
  filter(
    grepl(paste(commodity_vec, collapse = "|"), tolower(INSP_HOST)))

setwd(here::here('output'))
write.csv(aqas_f309_species_records,
          paste(
            'Pest interception records in F309 on',
            paste(commodity_vec,collapse = 'OR'), 
            'between', 
            begin_date, 
            end_date, 
            "in AQAS",
            ".csv",
          sep = " ")
)


```

# ARM data (the hard way)

```{r get records of the commodity coming in}
get.commodity = function(connection,
                         commodity
) {
  tbl(connection, sql( "SELECT
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_COMMODITY]")) |> 
    filter(COMMODITY_COMMON_NAME %in% commodity |
             COMMODITY_DISPLAY_NAME %in% commodity) |>
    dplyr::select(
      ID,
      INSPECTION_ID,
      DIAGNOSTIC_EVENT_ID,
      QUANTITY,
      QUANTITY_UNITS_NAME,
      COMMODITY_HOST_TYPE,
      DISPOSITION_CODE,
      CBP_COMMODITY_ID_NUMBER,
      INITIAL_COMMODITY_DISPLAY_NAME,
      INITIAL_COMMODITY_TAXONOMIC_DISPLAY_NAME,
      INITIAL_COMMODITY_COMMON_NAME,
      COMMODITY_DISPLAY_NAME,
      COMMODITY_TAXONOMIC_DISPLAY_NAME,
      COMMODITY_COMMON_NAME,
      TAXON_SIMPLE_NAME,
      PROPAGATIVE_MATERIAL_TYPE,
      FINAL_TAXON_SIMPLE_NAME,
      DESTINATION_CITY,
      SHIPMENT_IDENTIFIER_ID,
      DIAGNOSTIC_EVENT_ID,
      COUNTRY_OF_ORIGIN_ID, #these aren't complete records
      COUNTRY_OF_ORIGIN_NAME, #these aren't complete records
      QUARANTINE_RECOMMENDATION
    ) |> 
    collect() |> 
    rename('COMMODITY_ID' = 'ID') |> 
    mutate(across(starts_with("COMMODITY"), ~tolower(.)))
}

commod_records <- get.commodity(db_conn, commodity = commodity_vec )

```

get corresponding diagnostic results to those commodities

```{r}
get.dianostic.req = function(
    connection){
  tbl(connection, sql( "SELECT
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_REQUEST_ARTICLE]")) |> 
    dplyr::select(
      COMMODITY_ID,
      DIAGNOSTIC_REQUEST_ID,
      TREATMENT_SCHEDULE,
      REGULATORY_DETERMINATION_ID) |> 
    collect()
}

diag.req.records = get.dianostic.req(connection=db_conn
                                     ) |> filter(
                                       COMMODITY_ID %in% (
                                         commod_records |> pull(COMMODITY_ID)
                                         )
                                       )

#The BRG_DIAGNOSTIC_REQUEST_ARTICLE has COMMODITY_ID, but the FACT_DIAGNOSTIC_REQUEST does not for wahtever reason
## Should use the bridge to get to the FACT_INSPECTION's info about num dead & alive
get.dead.alive = function(
    connection){
  tbl(connection, sql( "SELECT
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_FACT_DIAGNOSTIC_REQUEST]")) |> 
    dplyr::select(
      ID,
      DIAGNOSTIC_REQUEST_NUMBER,
      NUMBER_ALIVE_ADULTS,
      NUMBER_ALIVE_CYSTS,
      NUMBER_ALIVE_EGGS,
      NUMBER_ALIVE_IMMATURE,
      NUMBER_ALIVE_PUPAE,
      NUMBER_DEAD_ADULTS,
      NUMBER_DEAD_CYSTS,
      NUMBER_DEAD_EGGS,
      NUMBER_DEAD_IMMATURE,
      NUMBER_DEAD_PUPAE,
      REMARKS
    ) |> 
    collect()
}

dead.alive.records = get.dead.alive(connection=db_conn) |> 
  filter(ID %in% (diag.req.records |> pull(DIAGNOSTIC_REQUEST_ID)))

#attach dead alive records from FACT_DR to BRG_DR
diag.req.records = diag.req.records |> 
  left_join(
    dead.alive.records,
    by = c('DIAGNOSTIC_REQUEST_ID'='ID')
    )

get.determinations = function(connection,
                              determ.ids) {
  tbl(connection, sql( "SELECT
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION]")) |> 
    #    filter(ID %in% determ.ids) |> 
    dplyr::select(
      ID,
      DIAGNOSTIC_REQUEST_ID,
      PEST_TAXON_ID,
      PEST_TAXON_SIMPLE_NAME,
      PEST_TAXONOMIC_NAME,
      QUARANTINE_STATUS_CONUS,
      QUARANTINE_STATUS_HAWAII,
      QUARANTINE_STATUS_PUERTO_RICO,
      ID_AUTHORITY,
      DETERMINATION_TYPE_ID,
      DETERMINED_BY_GROUP_ID,
      DETERMINATION_DATETIME
    ) |> 
    collect()
}

diag.determ.records = get.determinations(connection=db_conn#,
                                         # determ.ids = 
                                         #   (diag.req.records |> pull(REGULATORY_DETERMINATION_ID)
                                         #    )
) |> 
  filter(ID %in% (diag.req.records |> pull(REGULATORY_DETERMINATION_ID)))

#pull BRG_DIATNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON table 
## Just done to attach DIAGNOSTIC_NOT_POSSIBLE_FLAG in the event that a later molecular analysis was done but failed at a better determination (per Andy Carmichael's instructions, 8/12/24 on Teams)
get.diag.determ.not.possible = function(connection,
                                        determ.ids){
  tbl(connection, sql("SELECT
                      *
                      FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON]")) |> 
    # dplyr::filter(DIAGNOSTIC_DETERMINATION_ID %in% determ.ids) |> 
    collect() |> 
    dplyr::select(ID,
                  DIAGNOSTIC_DETERMINATION_ID,
                  DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON)
}

determ.not.possible.dat = get.diag.determ.not.possible(
  connection=db_conn#,
  # determ.ids = (diag.determ.records |> pull(ID))
) |>  filter(DIAGNOSTIC_DETERMINATION_ID %in% (diag.determ.records |> pull(ID)))

#merge these
diag.determ.records = diag.determ.records |> left_join(determ.not.possible.dat,
                                                       by=c('ID'='DIAGNOSTIC_DETERMINATION_ID'))

#parse the `final` determinations to the true, final determination
parse.final.determination = function(data) {
  data |> 
    group_by(DIAGNOSTIC_REQUEST_ID) |> 
    #Andy Carmichael says if that the most recent "DETERMINATION_CREATED_DATETIME" should be the most accurate, with very rare exceptions.  (The only time I can think of is if there was a molecular ID request, and it failed.)   In that case there DETERM_NOT_POSS_FLAG should be flagged.  
    #Filter for rows that are not a failed diagnostic determination
    dplyr::filter(is.na(DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON)) |> 
    #Filter the determination type - final has the highest TYPE_ID number
    dplyr::filter(DETERMINATION_TYPE_ID==max(DETERMINATION_TYPE_ID)) |> 
    #Filter for who ID'd it - the group_ID is a hierarchy of expertise, so take the max ID
    dplyr::filter(DETERMINED_BY_GROUP_ID == max(DETERMINED_BY_GROUP_ID)) |> 
    #The most recent is in practice the most correct - sometimes a final ID will be corrected at a later date, or someone will review for training or spot checking
    dplyr::filter(DETERMINATION_DATETIME == max(DETERMINATION_DATETIME))
}

diag.determ.records = diag.determ.records |> parse.final.determination()

#### Pull the pest taxonomy data from diagnostics ----
get.pest.taxonomy = function(connection,
                             pest.taxon.ids
) {
  tbl(connection, sql("SELECT
                       
                        *
                        FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_TAXONOMY]")) |> 
    #    filter(ID %in% pest.taxon.ids) |>  
    collect()
}

pest.taxonomy.records = get.pest.taxonomy(connection=db_conn) |> 
  filter(ID %in% (diag.determ.records |> pull(PEST_TAXON_ID)
                  )
         )

join.diagnostic.info = function(request,determ,taxonomy) {
  request |>
    left_join(determ,by=c('DIAGNOSTIC_REQUEST_ID'='DIAGNOSTIC_REQUEST_ID'),relationship='many-to-many')|> 
    left_join(taxonomy,by=c('PEST_TAXON_ID'='ID')) 
}

diag.info = join.diagnostic.info(request = diag.req.records,
                                 determ = diag.determ.records,
                                 taxonomy = pest.taxonomy.records) |> 
  mutate(COMMODITY_ID = as.character(COMMODITY_ID))

#use inner join to drop any commodity records that did not show up having a pest
complete.records = commod_records |> inner_join(diag.info,
                                                by='COMMODITY_ID') |> 
  filter(!is.na(PEST_TAXON_ID)) |> ungroup()
```

view results

```{r}
#number of commodity inspections with a diagnostic determination
complete.records

#times the commodity was inspected
commod_records
```

```{r save records}
setwd(here::here('output'))
write.csv(complete.records,
          paste(
            'Pest interception records on',
            paste(commodity_vec,collapse = 'OR'), 
            'between', 
            begin_date, 
            end_date, 
            ".csv",
          sep = " ")
)

write.csv(commod_records,
          paste(
            'Records of the commodities',
            paste(commodity_vec,collapse = 'OR'), 
            'being inspected between', 
            begin_date, 
            end_date, 
            ".csv",
          sep = " ")
)
```
